---
title: "RMarkdown Practice"
author: "Emily Huie"
date: "1/4/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


## Titles and Text
_EMILY_
**Emily**
###Emily
#EMily
Emily

## Links
[Emily BIS15 Repository](https://github.com/emhxm/BIS15W2022_ehuie/tree/main/lab1)

[Emily Huie](mailto:ezhuie@ucdavis.edu)
#InsertCodeChunks
```{r}
x=1
y=2
x+y
```
##Practice
1. Test out the arithmetic capabilities of R; experiment by doing addition, subtraction, multiplication, and division.
```{r}
(1+4)*36/2-12
```

2. Go back to your "RMarkdown Practice" file and experiment with titles, text, and syntax.
#TITLES
##Hello world. 
Hello world
_HELLOWORLD_
**HELLO WORLD**
# Links
[Emily BIS15 Repository](https://github.com/emhxm/BIS15W2022_ehuie/tree/main/lab1)

[Emily Huie](mailto:ezhuie@ucdavis.edu)
3. Copy and paste the following two pieces of code into the document (include the gray code "chunks").      

```{r}
#install.packages("tidyverse")
library("tidyverse")
```

```{r}
ggplot(mtcars, aes(x = factor(cyl))) +
    geom_bar()
```

```


```

